---
title: "R Tutorial: Precipitation Graphs, Pt 1"
author: "Nathan Byers"
date: "Tuesday, July 09, 2014"
output:
  html_document:
    theme: readable
---

--------------------------------------------------------------------------------

In this tutorial we will be using data from the `IDEMdata` package to manipulate
data sets, create custom graphs, and use functions and loops to automate the 
creation of these graphs. 

# The Data
To access the data, we need to download the package from a GitHub repository. The 
repository address is <a href="https://github.com/InDEM/IDEMdata" target="_blank">https://github.com/InDEM/IDEMdata</a>. To install it from GitHub, we use the `devtools` package.

```{r, message=FALSE}
library(devtools)
install_github("InDEM/IDEMdata")
```

Now we load the `IDEMdata` package and take a look at the package description
help file.

```{r, message=FALSE}
library(IDEMdata)
?IDEMdata
```

In the bottom right panel of <a href="rstudio.com" target="_blank">RStudio</a> 
you will see the file that describes what's in this package. At the very bottom 
of that file, click on "Index" and you will see all of the data sets that are in
the package. 

The "deep_river" data frames are data that were collected by sampling the 
Deep River - Portage Burns Watershed in Lake and Porter Counties. These data are
just a subset of sampling that took place in the watershed between April 2013
and March 2014. The purpose of the sampling was to get a baseline assessment of
the watershed for TMDL development and watershed planning. More information about
the study can be found <a href="http://www.in.gov/idem/nps/3893.htm" target="_blank">here</a>.

In part 1 of this tutorial we will be exploring these data sets
to become more familiar with them. In part 2 we will design graphs that visualize
the relationship between daily precipitation and the sampled data from the 
watershed. In part 2 we will write functions and use loops to automate the data 
manipulation necessary to make the graphs, and to automate the creation of the 
graphs themselves.